T-reX\@: Quantifying waste and material footprints in current and future \\ Life Cycle Assessment (LCA) databases

### 2025-02-09 00_T-reX_journal-upload-folder/T-reX_latex-files/cas-dc-template.tex
auto-update

%\usepackage[numbers]{natbib}
\usepackage[authoryear]{natbib}
% \usepackage[authoryear,longnamesfirst]{natbib}
%\usepackage{booktabs}
%\usepackage[tt=false]{libertine}
%\usepackage{xurl}

@@

	% Use if graphical abstract is present
	\begin{graphicalabstract}
		\includegraphics[width=\columnwidth]{grabs_T-reX_graphical-abstract.pdf}
		%\includegraphics[width=\columnwidth]{grabs_T-reX_graphical-abstract_MID-RES.png}
	\end{graphicalabstract}

@@

	\begin{keywords}
		circular economy, waste footprint, material footprint, life cycle assessment, critical raw materials, supply chain
	\end{keywords}

@@

	\begin{figure*}[]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX_method.pdf}
		\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}\label{fig:methods-flowchart}
	\end{figure*}

@@

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX_NMC811_WasteTotalSolid.pdf}
		\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser} showing the total solid waste footprint flows for the
			NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}\label{fig:sankey_waste}
	\end{figure}

@@

	\begin{figure*}[!htbp]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX-wastefootprint-totalANDcarbon.pdf}
		\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
	\end{figure*}

@@

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX-wastefootprint-processcontributions.pdf}
		\caption{Contribution by the `top-processes' to graphite demand inventory of the LiMn\(_2\)O\(_4\) battery from 2020 to 2100 under the SSP2 scenario using the baseline RCPs of the REMIND model}\label{fig:top_contribution}
	\end{figure}

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX-wastefootprint-sectorcontributions.pdf}
		\caption{Contribution of industrial sectors to the natural gas footprint of the LFP battery from 2020 to 2100 under the SSP2 scenario using the baseline RCP of the REMIND model}\label{fig:cpc_contribution}
	\end{figure}

@@

	\begin{figure*}[!htbp]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX-coalANDedip.pdf}
		\caption{Comparison of the (a) `T-reX – Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 – coal no longterm'. In the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
	\end{figure*}

@@

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX_credit_heatmap.pdf}
		\caption{CRediT authorship visualisation}\label{fig:credit_heatmap}
	\end{figure}

@@

	% Loading bibliography database
	\bibliography{01_T-reX_manuscript}

>>>>

%\usepackage[numbers]{natbib}
\usepackage[authoryear]{natbib}
% \usepackage[authoryear,longnamesfirst]{natbib}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
% \usepackage[tt=false]{libertine}
\usepackage{xcolor}
\usepackage{xurl}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\usepackage{float}
\usepackage{longtable}
\usepackage{geometry}
\usepackage{tabularx}
\usepackage{fontspec}
\setmainfont[Mapping=tex-text,Ligatures={Common,Rare,Discretionary}]{Linux Libertine}

@@

% Use if graphical abstract is present
\begin{graphicalabstract}
	\includegraphics[width=1\columnwidth]{figs/grabs_T-reX_graphical-abstract.pdf}
	%\includegraphics[width=\columnwidth]{figs/grabs_T-reX_graphical-abstract_MID-RES.png}
\end{graphicalabstract}

@@

\begin{keywords}
circular economy\sep waste footprint\sep material footprint\sep life cycle assessment\sep critical raw materials \sep supply chain
\end{keywords}

@@

\begin{figure*}[]
	\centering
	\includegraphics[width=1.8\columnwidth]{figs/T-reX_method.pdf}
	\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}\label{fig:methods-flowchart}
\end{figure*}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=0.8\columnwidth]{figs/T-reX_NMC811_WasteTotalSolid.pdf}
	\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser} showing the total solid waste footprint flows for the
	 NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}\label{fig:sankey_waste}
\end{figure}

@@

\begin{figure*}[!htbp]
	\centering
	\includegraphics[width=1.8\columnwidth]{figs/T-reX-wastefootprint-totalANDcarbon.pdf}
	\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
\end{figure*}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=0.8\columnwidth]{figs/T-reX-wastefootprint-processcontributions.pdf}
	\caption{Contribution by the `top-processes' to graphite demand inventory of the LiMn\(_2\)O\(_4\) battery from 2020 to 2100 under the SSP2 scenario using the baseline RCPs of the REMIND model}\label{fig:top_contribution}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{figs/T-reX-wastefootprint-sectorcontributions.pdf}
	\caption{Contribution of industrial sectors to the natural gas footprint of the LFP battery from 2020 to 2100 under the SSP2 scenario using the baseline RCP of the REMIND model}\label{fig:cpc_contribution}
\end{figure}

@@

\begin{figure*}[!htbp]
	\centering
	\includegraphics[width=1.8\columnwidth]{figs/T-reX-coalANDedip.pdf}
	\caption{Comparison of the (a) `T-reX – Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 – coal no longterm'. In the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
\end{figure*}

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=0.8\columnwidth]{figs/T-reX_credit_heatmap.pdf}
	\caption{CRediT authorship visualisation}\label{fig:credit_heatmap}
\end{figure}

@@

% Loading bibliography database
\bibliography{cas-refs}

### 2025-02-09 03_T-reX-latex_files-DRAFTING/.main.tex
auto-update

\begin{highlights}
    \item 'T-reX', a new open-source program for quantifying waste and material flows in Life Cycle Assessment (LCA)
    \item Assesses supply risks by calculating demand inventories for critical materials
    \item Simplifies quantification of user-specified waste and material footprint categories
    \item Rapidly identifies hotspots of waste generation and material demand in supply chains
    \item Presents a case study of five lithium-ion batteries to demonstrate T-reX's utility
\end{highlights}

@@

    \title{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases}
    \author[1]{Stewart Charles McDowall\corref{cor1}}
    \ead{s.c.mcdowall@cml.leidenuniv.nl}
    \author[1]{Elizabeth Lanphear}
    \author[1]{Stefano Cucurachi}
    \author[1]{Carlos Felipe Blanco}

@@

    \begin{abstract}
        % \cbox{Abstract word count: 150, Limit: 150}
        The quintessential principle of the `circular economy' is to minimise material consumption and waste generation. Thus, identifying and quantifying waste and material flows is critical. Life Cycle Assessment (LCA) is a powerful method, given its capacity to pinpoint hotspots of environmental impact throughout an activity's life cycle. Allowing the implementation of circular principles where could be most effective.

>>>>

\begin{highlights}
    \item ``T-reX'', a new open-source program for quantifying waste and material flows in Life Cycle Assessment (LCA)
    \item Assesses supply risks by calculating demand inventories for critical materials
    \item Simplifies quantification of user-specified waste and material footprint categories
    \item Rapidly identifies hotspots of waste generation and material demand in supply chains
    \item Presents a case study of five lithium-ion batteries to demonstrate T-reX's utility
\end{highlights}

@@

    \title{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases.\@}
    \author[1]{Stewart Charles McDowall\corref{cor1}}
    \ead{s.c.mcdowall@cml.leidenuniv.nl}
    \author[1]{Elizabeth Lanphear}
    \author[1]{Stefano Cucurachi}
    \author[1]{Carlos Felipe Blanco}

@@

    \begin{abstract}
        The quintessential principle of the `circular economy' is to minimise material consumption and waste generation. Thus, identifying and quantifying waste and material flows is critical. Life Cycle Assessment (LCA) is a powerful method, given its capacity to pinpoint hotspots of environmental impact throughout an activity's life cycle. Allowing the implementation of circular principles where could be most effective.

### 2025-02-09 03_T-reX-latex_files-DRAFTING/introduction.tex
auto-update

Life Cycle Assessment (LCA) is a useful method for the holistic estimation of the environmental impacts of products and processes. LCA can comprehensively evaluate impacts across the entire life cycle---from `cradle to grave'--- often identifying critical hotspots and guiding prioritisation of action~\citep{guinee2011lca}. The standard approach is to apply Life Cycle Impact Assessment (LCIA) methods (such as ReCiPe~\citep{huijbregts2016recipe} and CML~\citep{guinee2002cml}), which convert the inventory data into a set of impact scores based on the sum of the elementary flows (those between the technosphere and biosphere). These scores can then be aggregated into one metric for each impact category that can be compared across products and processes. Extending on standard LCA is \textit{ex-ante} LCA, which employs future scenarios to construct `prospective background databases' to predict the impacts of supply chains that have not yet been (and may never be) realised~\citep{cucurachi2018exante,blanco2020exante}.

@@

To facilitate the quantification of waste and material flows in LCA, we developed a Python package to extend the \texttt{brightway} open-source LCA framework~\citep{mutel2017brightway} and track these exchanges by translating them into inventory indicators and `pseudo' LCA impact (LCIA) categories. T-reX is `T(ool for) re-X' (reduce, reuse, recycle, etc...) and enables LCA practitioners to manipulate their databases to allow them to easily aggregate the mass and volume of any desired exchange, and to create flexible categories that differentiate between material categories, waste types, and treatment or EOL handling.

>>>>

Life Cycle Assessment (LCA) is a useful method for the holistic estimation of the environmental impacts of products and processes. LCA can comprehensively evaluate impacts across the entire life cycle---from `cradle to grave'---often identifying critical hotspots and guiding prioritisation of action~\citep{guinee2011lca}. The standard approach is to apply Life Cycle Impact Assessment (LCIA) methods (such as ReCiPe~\citep{huijbregts2016recipe} and CML~\citep{guinee2002cml}), which convert the inventory data into a set of impact scores based on the sum of the elementary flows (those between the technosphere and biosphere). These scores can then be aggregated into one metric for each impact category that can be compared across products and processes. Extending on standard LCA is \textit{ex-ante} LCA, which employs future scenarios to construct `prospective background databases' to predict the impacts of supply chains that have not yet been (and may never be) realised~\citep{cucurachi2018exante,blanco2020exante}.

@@

To facilitate the quantification of waste and material flows in LCA, we developed a Python package to extend the \texttt{brightway} open-source LCA framework~\citep{mutel2017brightway} and track these exchanges by translating them into inventory indicators and `pseudo' LCA impact (LCIA) categories. T-reX is `T(ool for) re-X' (reduce, reuse, recycle, etc\ldots) and enables LCA practitioners to manipulate their databases to allow them to easily aggregate the mass and volume of any desired exchange, and to create flexible categories that differentiate between material categories, waste types, and treatment or EOL handling.

### 2025-02-09 03_T-reX-latex_files-DRAFTING/methodology.tex
auto-update

In the default configuration of T-reX, there are 10 waste categories which are then further divided by their unit of measurement (kilograms and cubic meters) to create a total of 20 waste methods. The waste categories include incineration, recycling, and total waste, and are listed in the supplementary material in~\autoref{sec:supplementary}. One advantage of T-reX is that it can identify `hidden' waste exchanges that would otherwise be `consumed' by a treatment process and not leave the technosphere. Since `waste is not a service'~\citep{guinee2021wasteisnotaservice}, a characterisation factor of -1 is applied to the waste footprint methods (except CCS exchanges), changing the perspective from `waste consumed by treatment' to `waste generated by the activity'.

@@

To provide an example, we investigated five types of Li-ion batteries in the \texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3}, each represented by their unamended global market inventories:
\begin{itemize}[itemsep=0pt]
    \item Li-ion, LFP, rechargeable, prismatic
    \item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
    \item Li-ion, NMC111, rechargeable, prismatic
    \item Li-ion, NMC811, rechargeable, prismatic
    \item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\subsubsection{LCA calculations}
For each combination of activity, method, and database, a unified metric was calculated along with details of the top contributing processes. Additionally, for the T-reX methods, a contribution analysis was performed. Utilising the \texttt{bwa.compare\_activities\_by\_grouped\_leaves} function from the \texttt{brightway2\_analyzer} package~\citep{mutel2016brightway2analyzer}, a component of the \texttt{brightway} ecosystem, this function performs one subjective method of graph traversal on the impact matrix of the LCA object to a specified cutoff. The groups of the resulting leaves are collected by their Cooperative Patent Classification (CPC) codes [giving a representation of industry sector involvement]. T-reX is, an sich, agnostic to interests of its user, providing a tabulation of internal exchanges and an insight into the products and sectors in the supply chain of the activity that carry the most substantial responsibility for the final waste generation or material demand footprints.

>>>>

In the default configuration of T-reX, there are 10 waste categories which are then further divided by their unit of measurement (kilograms and cubic meters) to create a total of 20 waste methods. The waste categories include incineration, recycling, and total waste, and are listed in the supplementary material in~\autoref{sec:supplementary} [section 3 ]. One advantage of T-reX is that it can identify `hidden' waste exchanges that would otherwise be `consumed' by a treatment process and not leave the technosphere. Since `waste is not a service'~\citep{guinee2021wasteisnotaservice}, a characterisation factor of -1 is applied to the waste footprint methods (except CCS exchanges), changing the perspective from `waste consumed by treatment' to `waste generated by the activity'.

@@

To provide an example, we investigated five types of Li-ion batteries in the \texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3}, each represented by their unamended global market inventories. The system boundaies. These following markets were used for the example calculations and unaltered by the authors:
\begin{itemize}[itemsep=0pt]
    \item Li-ion, LFP, rechargeable, prismatic
    \item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
    \item Li-ion, NMC111, rechargeable, prismatic
    \item Li-ion, NMC811, rechargeable, prismatic
    \item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\subsubsection{LCA calculations}
For each combination of activity, method, and database, a unified metric was calculated along with details of the top contributing processes. Additionally, for the T-reX methods, a contribution analysis was performed. Utilising the \texttt{bwa.compare\_activities\_by\_grouped\_leaves} function from the \texttt{brightway2\_analyzer} package~\citep{mutel2016brightway2analyzer}, a component of the \texttt{brightway} ecosystem, this function performs one subjective method of graph traversal on the impact matrix of the LCA object to a specified cutoff. The groups of the resulting leaves are collected by their Cooperative Patent Classification (CPC) codes [giving a representation of industry sector involvement]. T-reX provides, essentially, a tabulation of internal exchanges and an insight into the products and sectors in the supply chain of the activity that carry the most substantial responsibility for the final waste generation or material demand footprints.
